Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
Apply some of Jonathan's feedback

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hierarchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to its counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representations are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction
of artifacts that are specific to theorem provers.

@@

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial as the reasoning principles we are allowed to use are directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
specified through parallel
reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contribution of this section is the definition of the logical relation as an
inductively defined relation and demonstration of how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function; that is, each type can
have at most one unique interpretation.

@@

In Section~\ref{sec:extension}, we extend the definition of the logical relation
to include terms with free variables, and prove that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst 2~\citep{autosubst2} for handling bindings and
CoqHammer~\citep{czajka2018hammer} for general-purpose automation.

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} \mathsf{id}_{\mathit{tm}}} \\
        [[idtm n]] & := & [[var n]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
        %
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        %
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation:}} ([[ a { rho } ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

We adapt from \citet{autosubst2} the notations for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$, the identity
substitution that maps each index to the corresponding term variable.  The
$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
$[[var Suc i]]$ to $[[rho i]]$.

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we first specify the equational theory
used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
relation is often called ``definitional equality'' in dependent type theories
because it defines the equivalence that the syntactic type system works
``up-to''.

@@

Our definition of convertibility for \lang{} is unusual in that it is directly
defined via parallel reduction, instead of using the related notion of
$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
language definition; a detailed argument of the equivalence between
$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{XXX}. \scw{Forward reference to further
  discussion, comparing with whnf reduction?}

@@

\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
System~\citep{barendregt1991introduction}, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion. Furthermore, discuss how this definition can be extended with
$\eta$-equivalence of functions in Section~\ref{XXX}. Finally, we compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

Although the use of de Bruijn indices complicates the variable rule, the rules
that involve binding, such as \rref{T-Pi,T-Abs} are straightforward.

@@

The elimination form for booleans, \rref{T-If} demonstrates dependent pattern
matching.  The result type of this expression $[[A {a}]]$, is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similar for the false
branch. As a result, the type system communicates the information gained from
the run-time test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is called a
\emph{motive} and has two free variables.  The substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching.  In the
typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
shifted to refer to the same binders.

>>>>

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we prove logical consistency, which states that empty types are not inhabited in an empty context, and demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal forms for well-typed open \emph{and}
closed terms. We include a full predicative universe hierarchy and
type-level computation to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so that
the material is accessible to readers with a general mathematical background.
That said, our explanations do not stray too far away from our proof scripts.
We link each lemma directly to its counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
The typeset proofs purposefully follow the
mechanized proofs closely while avoiding, as much as possible, artifacts specific to Coq.

@@

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial, as the reasoning principles we are allowed to use are directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
specified by parallel
reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contributions of this section are the definition of the logical relation as an
inductively defined relation and demonstrating how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function: that is, each type can
have at most one unique interpretation.

@@

In Section~\ref{sec:extension}, we extend the definition of the logical relation
to include terms with free variables, and prove that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$ law
into the conversion rule for \lang{}.

In Section~\ref{sec:logrelmech}, we discuss the details of our
mechanization, including our use of existing libraries such as
Autosubst 2~\citep{autosubst2} for handling bindings and
CoqHammer~\citep{czajka2018hammer} for general-purpose automation.

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
        [[idtm n]] & := & [[var n]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
        %
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        %
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

We adapt from \citet{autosubst2} notation for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$, the identity
substitution that maps each index to the corresponding term variable.  The
$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
$[[var Suc i]]$ to $[[rho i]]$.

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we first specify the equational theory
used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
relation is often called \textit{definitional equality} in dependent type theories
because it defines the equivalence that the syntactic type system works
up to.

@@

Our definition of convertibility for \lang{} is unusual in that it is directly
defined via parallel reduction, instead of using the related notion of
$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
language definition; a detailed argument of the equivalence between
$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{sec:discuss}. \scw{Forward reference to further
  discussion, comparing with whnf reduction?}

@@

\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
System~\citep{barendregt1991introduction}, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta}
% Furthermore, discuss how this definition can be extended with
% $\eta$-equivalence of functions in Section~\ref{XXX}
Finally, we compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

Although the use of de Bruijn indices complicates the variable rule, the rules
that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.

@@

The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression $[[A {a}]]$ is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the % run-time \yl{commented out run-time per Jonathan's comment}
test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is the motive and has two free variables.  The substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching.  In the
typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
shifted to refer to the same binders.
